---
title: "PAR Position"
author: "Mathieu Adgé"
date: "2026-08-07"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls())
library(tidyverse)
library(nlme)
library(MuMIn)
library(patchwork)
library(emmeans)
library(performance)
```


```{r import data}
data <- read.csv("../data/Light_position.csv") %>% 
  filter(Sat_2==0) %>% 
   mutate(across(c(Zone,Position,Orientation), as.factor), datetime=ymd_hms(datetime), hm=hour(datetime) + minute(datetime) / 60, Date = as.Date(day)) 
```

We study the SH81 platform to see if the extreme value obtained for position (4.5,4.5) is common with values in other positions under the plateform 

```{r model}
data_Z3<- data %>% 
  filter(Zone==3 & Orientation == "Haut") %>% 
  mutate(facteur=log(facteur))

M2 <- gls(
  facteur ~ Position,
  correlation = corCAR1(form = ~ hm|Position),
  data = data_Z3,
  method = "REML"
)
```

```{r model info}
summary(M2)

acf(resid(M2, type="normalized"), lag.max = 36, main = "ACF")
```

```{r visual check on model residuals}
res <- residuals(M2, type="normalized")
    fitted <- fitted(M2)
  resfit <- data_Z3 %>% 
    mutate(res_log= res , fitted_log=fitted)
  
           p_hist <-    ggplot(resfit, aes(x = res_log)) +
                geom_histogram(bins = 50,
                               fill = "lightblue", color = "black")+
                labs(title = "")
           p_qq <-    ggplot(resfit, aes(sample = res_log)) +
                stat_qq() +
                stat_qq_line()+
                labs(title = "Normal Q-Q Plot")
           p_fit <-   ggplot(resfit, aes(y = res_log, x = fitted_log)) +
                geom_point()+
                labs(title = "")
           p_res_pos <-     ggplot(resfit, aes(y = res_log, x = Position, fill= Position)) +
                geom_boxplot()
              
     p_diag <- (p_hist | p_qq) / (p_fit | p_res_pos)
p_diag

```

```{r}
emm_Position    <- emmeans(M2,~ Position, mode = "appx-satterthwaite")
print(emm_Position)

emm_table <- as.data.frame(emm_Position)


facteur_mult_tb = emm_table %>% 
  mutate(t = -(lower.CL-emmean)/SE) %>% 
  summarise(t_mean= mean(t)) 

facteur_mult = facteur_mult_tb[[1,1]]

ER_max = 0.13
propagation = sqrt(2*(ER_max^2))

emm_table_norm <- as.data.frame(emm_table) %>% 
    mutate(across(c(emmean,SE), ~ exp(.))) %>% 
  mutate(lower.CL = emmean-facteur_mult*sqrt((SE^2)+(propagation^2)), 
         upper.CL = emmean+facteur_mult*sqrt((SE^2)+(propagation^2))) %>% 
  add_row(Position = as.factor(4.5), emmean = 18, SE =1.01, df =NA , lower.CL = 15.98 , upper.CL = 20.02)

### Fig 4. Marginal mean attenuation coefficient of direct light under SH81 compared to the marginal mean attenuation coefficient at different positon 


emm_table_norm_ref  <- emm_table_norm %>% filter(Position == 4.5)
emm_table_norm_spat <- emm_table_norm %>% filter(Position != 4.5)


emm_table_norm_spat <- emm_table_norm_spat %>%
  mutate(Position = factor(Position, levels = sort(unique(Position))))


ggplot() +


  annotate(
    "rect",
    xmin = -Inf, xmax = Inf,
    ymin = emm_table_norm_ref$lower.CL,
    ymax = emm_table_norm_ref$upper.CL,
    fill = "#DCB0F2FF",
    alpha = 0.5
  ) +


  geom_point(
    data = emm_table_norm_spat,
    aes(x = Position, y = emmean),
    size = 3
  ) +


  geom_errorbar(
    data = emm_table_norm_spat,
    aes(x = Position, ymin = lower.CL, ymax = upper.CL),
    width = 0.15,
    linewidth = 0.8
  ) +

  labs(
    x = "Position",
    y = "Estimated marginal mean",
    caption = "The rectangle represents the 95% CI for position 4.5 in direct light factor attenuation timeserie model"
  ) +

  theme_minimal() +
  theme(
        legend.text = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.spacing = unit(0.5, "lines"),
    legend.position = "bottom",
    plot.caption = element_text(hjust = 0.5)
  )
```
We study the SH81-UV platform to see if the different values obtained for position (3,3) and (4.5,4.5) is common with values in other positions under the plateform 


```{r data selection}
data_Z5<- data %>% 
  filter(Zone==5 & Orientation == "Haut") %>% 
  mutate(facteur=log(facteur))
```

```{r model selection}
M3 <- gls(
  facteur~Position,
  weights = varIdent(form=~1|Position),
  data = data_Z5,
  method = "REML"
)
```

```{r model info}
summary(M3) 

acf(resid(M3, type="normalized"), lag.max = 36, main = "ACF")

r.squaredLR(M3)
```

```{r visual check on model residuals}
res <- residuals(M3, type="normalized")
    fitted <- fitted(M3)
  resfit <- data_Z5 %>% 
    mutate(res_log= res , fitted_log=fitted)
  
           p_hist <-    ggplot(resfit, aes(x = res_log)) +
                geom_histogram(bins = 50,
                               fill = "lightblue", color = "black")+
                labs(title = "Histogramme des residus")
           p_qq <-    ggplot(resfit, aes(sample = res_log)) +
                stat_qq() +
                stat_qq_line()+
                labs(title = "Normal Q-Q Plot")
           p_fit <-   ggplot(resfit, aes(y = res_log, x = fitted_log)) +
                geom_point()+
                labs(title = "Valeurs ajustees")
           p_res_pos <-     ggplot(resfit, aes(y = res_log, x = Position, fill= Position)) +
                geom_boxplot()
              
     p_diag <- (p_hist | p_qq) / (p_fit | p_res_pos)
p_diag
```

```{r}
emm_Position    <- emmeans(M3,~ Position, mode = "appx-satterthwaite")
print(emm_Position)

con_Position <- contrast(emm_Position, method = "pairwise" ,adjust = "tukey")
print(con_Position)

ct_ci <- as.data.frame(confint(con_Position, level=0.95, adjust="tukey"))

ER_max = 0.13
propagation = sqrt(2*(ER_max^2))


con_Pos_table <- as.data.frame(con_Position) %>% 
  left_join(ct_ci) %>% 
  mutate(
    phys_detectable = (lower.CL > propagation) | (upper.CL < -propagation),
    sign_capteur = if_else(
      phys_detectable,
"Difference greater than the instrumental error",
      "Difference inferior than the instrumental error"
    ),
    filtre_sign = as.integer(phys_detectable)
  )

df <- con_Pos_table %>% 
  separate(contrast, into = c("Position1", "Position2"), sep = "\\s-\\s") %>% 
    mutate(
    Position1_num = as.numeric(str_extract(Position1, "\\d+")),
    Position2_num = as.numeric(str_extract(Position2, "\\d+")),
    p_num = as.numeric(`p.value`),
    p_cat = case_when(
      p_num < 0.001 ~ "p < 0.001",
      p_num < 0.01  ~ "p < 0.01",
      p_num < 0.05  ~ "p < 0.05",
      TRUE          ~ "Statistically non significant"
    ),
    phys = if ("phys_detectable" %in% names(.)) phys_detectable else filtre_sign == 1,)

    
position <- sort(unique(c(con_Pos_table$Position1, df$Position2)))

df$p_cat <- factor(
  df$p_cat,
  levels = c("p < 0.001", "p < 0.01", "p < 0.05", "Statistically non significant")
)



### Fig.5 Results of pairwise post-hoc comparisons of direct light attenuation at different positions for the SH81-UV platform 
ggplot(df, aes(x = factor(Position1_num), y = factor(Position2_num))) +
  geom_tile(aes(fill = p_cat), color = "white", linewidth = 0.25) +
  geom_point(
    data = df %>% filter(phys),
    shape = 4, size = 2, stroke = 1
  ) +
  scale_fill_manual(
    name = "p-value",
    values = c(
      "p < 0.001" = "#8B0000",
      "p < 0.01"  = "#E41A1C",
      "p < 0.05"  = "#FDB863",
      "Statistically non significant" = "grey85"
    ),
    na.value = "grey95"
  ) +
  coord_equal() +
  labs(
    x = NULL, y = NULL,
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"), 
  )
```
Reflected light analysis 

```{r}
### SH51
data_Z1_ref<- data %>% 
  filter(Zone==1 & Orientation == "Bas") %>% 
  mutate(facteur=log(facteur)) %>% 
  filter(PAR!=12)
```

```{r}
f1 <- formula(facteur~ Position)
M1<-gls(facteur~Position,method="REML", data = data_Z1_ref)
```


```{r}
res <- residuals(M1, type="normalized")
    fitted <- fitted(M1)
  resfit <- data_Z1_ref %>% 
    mutate(res_log= res , fitted_log=fitted)
  
           p_hist <-    ggplot(resfit, aes(x = res_log)) +
                geom_histogram(bins = 50,
                               fill = "lightblue", color = "black")+
                labs(title = "Histogramme des residus")
           p_qq <-    ggplot(resfit, aes(sample = res_log)) +
                stat_qq() +
                stat_qq_line()+
                labs(title = "Normal Q-Q Plot")
           p_fit <-   ggplot(resfit, aes(y = res_log, x = fitted_log)) +
                geom_point()+
                labs(title = "Valeurs ajustees")
           p_res_pos <-     ggplot(resfit, aes(y = res_log, x = Position, fill= Position)) +
                geom_boxplot()
              
     p_diag <- (p_hist | p_qq) / (p_fit | p_res_pos)
p_diag
```

```{r}
summary(M1)

r.squaredLR(M1)
```

```{r}
emm_Position    <- emmeans(M1,~ Position, mode = "appx-satterthwaite")

con_Position <- contrast(emm_Position, method = "pairwise" ,adjust = "tukey")

ct_ci <- as.data.frame(confint(con_Position, level=0.95, adjust="tukey"))

ER_max = 0.13
propagation = sqrt(2*(ER_max^2))


con_Pos_table_Z1 <- as.data.frame(con_Position) %>% 
  left_join(ct_ci) %>% 
  mutate(
    phys_detectable = (lower.CL > propagation) | (upper.CL < -propagation),
    sign_capteur = if_else(
      phys_detectable,
"Difference greater than the instrumental error",
      "Difference inferior than the instrumental error"
    ),
    filtre_sign = as.integer(phys_detectable)
  ) %>% 
  mutate(Zone=1)

```
Aucune position n'a une différence significative stastiquement et physiquement

### Z3 

```{r}
data_Z3_ref<- data %>% 
  filter(Zone==3 & Orientation == "Bas") %>% 
  mutate(facteur=log(facteur)) %>% 
  filter(PAR !=12)
```

```{r}
M2 <- gls(
  facteur ~ Position,
  correlation = corCAR1(form = ~ hm|Position),
  data = data_Z3_ref,
  method = "REML"
)
```

```{r}
res <- residuals(M2, type="normalized")
    fitted <- fitted(M2)
  resfit <- data_Z3_ref %>% 
    mutate(res_log= res , fitted_log=fitted)
  
           p_hist <-    ggplot(resfit, aes(x = res_log)) +
                geom_histogram(bins = 50,
                               fill = "lightblue", color = "black")+
                labs(title = "Histogramme des residus")
           p_qq <-    ggplot(resfit, aes(sample = res_log)) +
                stat_qq() +
                stat_qq_line()+
                labs(title = "Normal Q-Q Plot")
           p_fit <-   ggplot(resfit, aes(y = res_log, x = fitted_log)) +
                geom_point()+
                labs(title = "Valeurs ajustees")
           p_res_pos <-     ggplot(resfit, aes(y = res_log, x = Position, fill= Position)) +
                geom_boxplot()
              
     p_diag <- (p_hist | p_qq) / (p_fit | p_res_pos)
p_diag
```

```{r}
summary(M2)

r.squaredLR(M2)
```

```{r}
emm_Position    <- emmeans(M2,~ Position, mode = "appx-satterthwaite")

con_Position <- contrast(emm_Position, method = "pairwise" ,adjust = "tukey")

ct_ci <- as.data.frame(confint(con_Position, level=0.95, adjust="tukey"))

ER_max = 0.13
propagation = sqrt(2*(ER_max^2))


con_Pos_table_Z3 <- as.data.frame(con_Position) %>% 
  left_join(ct_ci) %>% 
  mutate(
    phys_detectable = (lower.CL > propagation) | (upper.CL < -propagation),
    sign_capteur = if_else(
      phys_detectable,
"Difference greater than the instrumental error",
      "Difference inferior than the instrumental error"
    ),
    filtre_sign = as.integer(phys_detectable),
    Zone = 3
  )

```

### Z4 

```{r}
data_Z4_ref<- data %>% 
  filter(Zone==4 & Orientation == "Bas") %>% 
  mutate(facteur=log(facteur)) %>% 
  filter(PAR !=12)

```


```{r}
M3 <- gls(
  facteur~Position,
  weights = varIdent(form=~1|Position),
  data = data_Z4_ref,
  method = "REML"
)

summary(M3)

r.squaredLR(M3)
```

```{r}
res <- residuals(M3, type="normalized")
    fitted <- fitted(M3)
  resfit <- data_Z4_ref %>% 
    mutate(res_log= res , fitted_log=fitted)
  
           p_hist <-    ggplot(resfit, aes(x = res_log)) +
                geom_histogram(bins = 50,
                               fill = "lightblue", color = "black")+
                labs(title = "Histogramme des residus")
           p_qq <-    ggplot(resfit, aes(sample = res_log)) +
                stat_qq() +
                stat_qq_line()+
                labs(title = "Normal Q-Q Plot")
           p_fit <-   ggplot(resfit, aes(y = res_log, x = fitted_log)) +
                geom_point()+
                labs(title = "Valeurs ajustees")
           p_res_pos <-     ggplot(resfit, aes(y = res_log, x = Position, fill= Position)) +
                geom_boxplot()
              
     p_diag <- (p_hist | p_qq) / (p_fit | p_res_pos)
p_diag
```

```{r}
emm_Position    <- emmeans(M3,~ Position, mode = "appx-satterthwaite")

con_Position <- contrast(emm_Position, method = "pairwise" ,adjust = "tukey")

ct_ci <- as.data.frame(confint(con_Position, level=0.95, adjust="tukey"))

ER_max = 0.13
propagation = sqrt(2*(ER_max^2))


con_Pos_table_Z4 <- as.data.frame(con_Position) %>% 
  left_join(ct_ci) %>% 
  mutate(
    phys_detectable = (lower.CL > propagation) | (upper.CL < -propagation),
    sign_capteur = if_else(
      phys_detectable,
"Difference greater than the instrumental error",
      "Difference inferior than the instrumental error"
    ),
    filtre_sign = as.integer(phys_detectable),
    Zone=4
  )

```

### Z5 

```{r}
data_Z5_ref<- data %>% 
  filter(Zone==5 & Orientation == "Bas") %>% 
  mutate(facteur=log(facteur)) %>% 
  filter(!PAR %in% c(5,12))

M1<-gls(facteur~Position,method="REML", data = data_Z5_ref)
```




```{r}
res <- residuals(M1, type="normalized")
    fitted <- fitted(M1)
  resfit <- data_Z5_ref %>% 
    mutate(res_log= res , fitted_log=fitted)
  
           p_hist <-    ggplot(resfit, aes(x = res_log)) +
                geom_histogram(bins = 50,
                               fill = "lightblue", color = "black")+
                labs(title = "Histogramme des residus")
           p_qq <-    ggplot(resfit, aes(sample = res_log)) +
                stat_qq() +
                stat_qq_line()+
                labs(title = "Normal Q-Q Plot")
           p_fit <-   ggplot(resfit, aes(y = res_log, x = fitted_log)) +
                geom_point()+
                labs(title = "Valeurs ajustees")
           p_res_pos <-     ggplot(resfit, aes(y = res_log, x = Position, fill= Position)) +
                geom_boxplot()
              
     p_diag <- (p_hist | p_qq) / (p_fit | p_res_pos)
p_diag
```

```{r}
summary(M1)

r.squaredLR(M1)
```

```{r}
emm_Position    <- emmeans(M1,~ Position, mode = "appx-satterthwaite")

con_Position <- contrast(emm_Position, method = "pairwise" ,adjust = "tukey")

ct_ci <- as.data.frame(confint(con_Position, level=0.95, adjust="tukey"))

ER_max = 0.13
propagation = sqrt(2*(ER_max^2))


con_Pos_table_Z5 <- as.data.frame(con_Position) %>% 
  left_join(ct_ci) %>% 
  mutate(
    phys_detectable = (lower.CL > propagation) | (upper.CL < -propagation),
    sign_capteur = if_else(
      phys_detectable,
"Difference greater than the instrumental error",
      "Difference inferior than the instrumental error"
    ),
    filtre_sign = as.integer(phys_detectable), 
    Zone=5
  )

```

```{r}
### Fig.7 Results of pairwise post-hoc comparisons of reflected light attenuation factor

plot_signi = function(table, r_squared, freedom, phi=NA){
  zone_labels <- c(
  "1" = "Shade 51%",
  "4" = "Shade 51% - UV",
  "3" = "Shade 81%",
  "5" = "Shade 81% - UV"
)
  df <- table %>% 
  separate(contrast, into = c("Position1", "Position2"), sep = "\\s-\\s") %>% 
    mutate(
    Position1_num = as.numeric(str_extract(Position1, "\\d+")),
    Position2_num = as.numeric(str_extract(Position2, "\\d+")),
    p_num = as.numeric(`p.value`),
    p_cat = case_when(
      p_num < 0.001 ~ "p < 0.001",
      p_num < 0.01  ~ "p < 0.01",
      p_num < 0.05  ~ "p < 0.05",
      TRUE          ~ "Statistically non significant"
    ),
    phys = if ("phys_detectable" %in% names(.)) phys_detectable else filtre_sign == 1,
    )

position <- sort(unique(c(df$Position1, df$Position2)))



df$p_cat <- factor(
  df$p_cat,
  levels = c("p < 0.001", "p < 0.01", "p < 0.05", "Statistically non significant")
)

if(is.na(phi)==T){
  caption = paste("R² =", r_squared, "df =", freedom)
}else{
  caption = paste("R² =", r_squared, "df =", freedom, "Φ =", phi)
}


plot <- ggplot(df, aes(x = factor(Position1_num), y = factor(Position2_num))) +
  geom_tile(aes(fill = p_cat), color = "white", linewidth = 0.25) +
  geom_point(
    data = df %>% filter(phys),
    shape = 4, size = 2, stroke = 1
  ) +
  scale_fill_manual(
    name = "p-value",
    values = c(
      "p < 0.001" = "#8B0000",
      "p < 0.01"  = "#E41A1C",
      "p < 0.05"  = "#FDB863",
      "Statistically non significant" = "grey85"
    ),
    na.value = "grey95"
  ) +
  facet_wrap(~Zone, labeller= labeller(Zone=zone_labels)) +
  coord_equal() +
  labs(
    x = NULL, y = NULL,
    caption = caption
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"), 
    plot.caption = element_text(hjust = 0.5)
  )

return(plot)
}
```

```{r}
p1 <- plot_signi(con_Pos_table_Z1, r_squared = 0.13, freedom = 162)+
  theme(legend.position = "none")
p3  <-plot_signi(con_Pos_table_Z3, r_squared = 0.78, phi = 0.38, freedom =192)+
theme(legend.position = "none")
p4 <- plot_signi(con_Pos_table_Z4, r_squared = 0.37, freedom = 218)
p5 <- plot_signi(con_Pos_table_Z5, r_squared = 0.27, freedom = 160)+
  theme(legend.position = "none")

     p <- (p1 | p4) / (p3 | p5) + 
       plot_layout(guides = "collect")
p

```

